---
title: "Case Study - 1"
date: "November 8, 2016"
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,                         output_file='papers/Report.html') } )
output:
  html_document:
    keep_md: true
---

#Introduction 

The Organization for Economic Co-operation and Development (OECD) is a global organization consisting of representative nations from North and South America, Europe and Asia.  Member nations contribute time and resources to the organization's efforts to continually improve processes for fostering prosperity and fighting poverty through economic growth and financial stability.  Analysis of GDP data for OECD versus non-OECD countries with high income designations is of particular interest in this study.  GDP data for 190 ranked countries and the related educational data will be analyzed for this purpose.  In a second part of the study using the same data, a perceptual analysis of the relationship between GDP and income group will be performed by generating a graphical representation of GDP by income group.  The related quantile data will also be included in a tabular view for analysis to determine the number of countries with lower middle income that are among the 38 nations with the highest GDP.


```{r, include=FALSE}
# This Code sets up the environment by running the analysis/SetupEnvironment.R Script.  Output is not included in the RMarkdown output.
knitr::opts_chunk$set(echo = TRUE)
source("analysis/SetupEnvironment.R")
```

Data source - https://d396qusza40orc.cloudfront.net. 

## Download and Tidy GDP Data
GDP Data is downloaded and made tidy by running the script analysis/gdp_data.r which does the following:

1. Raw GDP data - downloaded from https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv to analysis/data/raw_fgdp.csv.

2. Rows with invalid observational data are removed from the dataframe using dataframe subset notation.

3. Unnecessary Columns are removed. The only columns needed for the analysis are those representative of the CountryCode, Ranking and GDP. These columns are renamed using the names function.

4. Invalid numeric characters are removed from the GDP column using the gsub function so that the column can be converted to type numeric.

5. The cleaned data set is written out to "analysis/data/tidy_gdp.csv" for later inspection.

```{r}
source("Analysis/gdp_data.R",echo = TRUE )
```

## Download and Tidy Education Data
Education Data is downloaded and cleaned by running the script analysis/educ_data.r which does the following:

1. Raw education data is downloaded from https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv.

2. Unnecessary Columns are removed from the dataframe using dataframe subset notation

3. The tidy data set is written out to "analysis/data/tidy_educ.csv" for later inspection.

```{r}
source("Analysis/educ_data.R",echo = TRUE)
```

## Merge GDP and Educational Data
The GDP and educational datasets are now merged together by running the script analysis/MergeData.r which does the following:

1. The GDP and Education Dataframes are merged together using the merge function and associated by CountryCode variable.  

```{r}
source("Analysis/MergeData.R",echo = TRUE)
```
One important note about the merge process, there was an entry from the GDP data set for (SDS) that didn't have a matching country code in the education dataset.  This produced a single line for this country that included NAs for educational data and was consequently removed during the merge operation.

## Order GDP and Educational Data
Arrange function is called to order the dataframes
The merged dataset is now ordered by running the script analysis/OrderData.r which does the following:

1. GDP and Education Dataframes are ordered by descending GDP value.

```{r}
source("Analysis/OrderData.R",echo = TRUE)
```

# Analysis Questions

####Match the data based on the country shortcode. How many of the IDs match?

```{r}
message("No. of ID's matched between the two data sets: ",length(merge_data))
```


#### Sort the data frame in ascending order by GDP rank. What is the 13th country in the resulting data frame?

```{r}
merge_data[13,]
```

St. Kitts and Nevis holds the 13th spot.


#### What are the average GDP rankings for the "High income: OECD" and "High income:nonOECD" groups?

The following code calculates the average GDP Rankings for the High Income OECD and non-OECD groups.

```{r}
# Get GDP Rankings for High Income OECD Group
highincome.oecd <- merge_data$Ranking[merge_data$Income.Group == "High income: OECD"]
highincome.oecd <- highincome.oecd[!is.na(highincome.oecd)]
highincome.oecd.average <- mean(highincome.oecd)
print(highincome.oecd.average)

# Get GDP Rankings for High Income non-OECD Group
highincome.nonoecd <- merge_data$Ranking[merge_data$Income.Group == "High income: nonOECD"]
highincome.nonoecd <- highincome.nonoecd[!is.na(highincome.nonoecd)]
highincome.nonoecd.average <- mean(highincome.nonoecd)
print(highincome.nonoecd.average)
```

The average GDP rankings for the OECD high income member countries is significantly lower than for non-OECD high income countries.  

#### Plot the GDP for all of the countries by Income Group

The following boxplot shows the GDP data by income groups for all countries.

```{r}
# Get GDP Rankings for High Income OECD Group
ggplot(merge_data, aes(Income.Group, GDP, color = Income.Group)) + 
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Observations from the boxplot - 

1. High income OECD has more varition in the GDP income values compared to rest of the groups

2. There seems to be a difference in the mean of value of GDP income between different groups of income



#### Cut the GDP ranking into 5 separate quantile groups. How many countries are Lower middle income but among the 38 nations with highest GDP?

The following code splits the GDP Rankings into 5 quantile groups and creates a table of these quartiles versus Income Groups.

```{r}
# Get GDP Rankings for High Income OECD Group
merge_data$quartile <- dplyr::ntile(merge_data$Ranking, 5) 
merge_data[merge_data$quartile == 1 & merge_data$Income.Group == "Lower middle income",]
table(merge_data$quartile,merge_data$Income.Group)
```


The table shows that there are 5 countries with the lower middle income designation that have GDP rankings in the top 38.

# Conclusion

The results of study showed a much higher Mean GDP ranking for countries with high income OECD designations than countries with high income designations that didn't participate in OECD.  This provides evidence that there may be a relationship between participation in the program and the size of a country's GDP.  Futher analysis should be conducted to better understand the relationship.  In the second part of the study it was shown that a small minority of countries with lower middle income designations were in the top quantile of GDP rankings.  This provides evidence that countries with lower middle income may have higher GDP value/rank.